An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-03 EconGrowthSeminar.Rnw
Many editing changes in preparation for seminar on 4 April 2013.

  Cobb-Douglas: \cdEqn
  \vspace{\baselineskip}

  Cobb-Douglas with Energy: \cdeEqn
  \vspace{\baselineskip}

  Constant Elasticity of Substitution:
  \cesEqn
  \vspace{\baselineskip}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, factor of production ($k$, $l$, $q$, $x$, or $u$)\\
      \hline
      \lambdaRow\\
      $m$ & ``influence'' of factor $f$ ($\alpha$, $\beta$, or $\gamma$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      \lambdaRow\\
      \alphaRow\\
      \betaRow\\
      \gammaRow\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      $\gamma$   &   fitting coefficient\\
      \lambdaRow\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``influence'' of ($k$) vs. ($l$)\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``influence'' of ($kl$) vs. ($e$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      $a_0$ & fitting parameter\\
      $c_t$ & fitting parameter\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

    \item{Capital stock ($K$): market value of productive goods (e.g., machines, tools, infrastructure)}
    \vspace{\baselineskip}

    \item{Labor ($L$): human work in an economy (people or hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total employment} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

    \item{Useful work ($U$): energy that generates economic activity directly}
  \end{itemize}
  \begin{center}Note: $q$ and $x$ are identical unless energy mix changes significantly \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Nat. gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{4}{*}{FAOStat} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{4}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
                              &                          & (1980-2011: Developed; &                         \\
                              &                          & 1991-2011: Other) &                         \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors. &  \multirow{2}{*}{\citet{Warr:2006hc}}\\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model Equation} %----------------------------
    \begin{center} \sfEqn \end{center}
    \vspace{\baselineskip}

    \begin{itemize}
      \item{Illustrative}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Irrepressible ``progress'' of technology}
        \item{Independent of factor of production ($f$)}
      \end{itemize}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

\begin{frame}{Example: Single-factor model with $q$} %----------------------------
  \begin{center}$y = \mathrm{e}^{\lambda (t-t_0)}q^{\gamma}$\end{center}
  \begin{itemize}
    \item{Select $\lambda$, $\gamma$}
    \item{To minimize \\ $SSE = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on Cobb-Douglas Model Equation} %----------------------------
  \begin{center} \cdeEqn \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$k$, $l$, and $e$ factors of production ``blend'' with $\alpha + \beta + \gamma = 1$}
    \item{Example: \citet{Froling:2009vo}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \cesEqn \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{Example: \citet{Stern:2012ey}}
    \item{$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, $\rho$ estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Derived from thermodynamics \citep{Kummel:1980wx}}
    \item{Time-dependent technology $\rightarrow$ efficiency ratios}
    \item{Productivity $\uparrow$ if}
    \begin{itemize}
      \item{Efficiency of producing capital stock increases ($k$ > $l + e$)}
      \item{Energy efficiency of labor increases ($l$ > $e$)}
    \end{itemize}
    \item{Example: \citet{Warr:2006hc}}
    \item{$a_0$ and $c_t$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

>>>>

  Cobb-Douglas (CD): \cdEqn
  \vspace{\baselineskip}

  Cobb-Douglas (CD) with Energy: \cdeEqn
  \vspace{\baselineskip}

  Constant Elasticity of Substitution (CES):
  \cesEqn
  \vspace{\baselineskip}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, factor of production ($k$, $l$, $q$, $x$, or $u$)\\
      \hline
      \lambdaRow\\
      $m$ & ``influence'' of factor $f$ ($\alpha$, $\beta$, or $\gamma$)\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

    Example: \citet{Nel:2009wq} use $f$ = $q$
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      \lambdaRow\\
      \alphaRow\\
      \betaRow\\
      \gammaRow\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

    Example: \citet{Froling:2009vo}
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      $\gamma$   &   fitting coefficient\\
      \lambdaRow\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``influence'' of ($k$) vs. ($l$)\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``influence'' of ($kl$) vs. ($e$)\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

    Example: \citet{Stern:2012ey}
  \end{center}
\end{frame}

@@

    \begin{tabular}{|c|l|}\hline
      $a_0$ & fitting parameter\\
      $c_t$ & fitting parameter\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

    Example: \citet{Warr:2006hc}
  \end{center}
\end{frame}

@@

    \item{Capital stock ($K$): market value (real 2005\$) of productive goods (e.g., machines, tools, infrastructure)}
    \vspace{\baselineskip}

    \item{Labor ($L$): human work in an economy (persons or hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total employment (persons)} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

    \item{Useful work ($U$): energy that generates economic activity directly \citep{Warr:2006hc}}
  \end{itemize}
  \begin{center}Note: $q$ and $x$ are identical unless energy mix changes significantly \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Natural gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{2}{*}{FAOStat} & \multirow{2}{*}{\includegraphics[scale=0.06]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{2}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors. &  \multirow{2}{*}{\citet{Warr:2006hc}}\\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Example: Single-factor model with $q$} %----------------------------
  \begin{center}$y = A \: q^{\gamma}; A \equiv \mathrm{e}^{\lambda (t-t_0)}$\end{center}
  \begin{itemize}
    \item{Select $\lambda$, $\gamma$ to minimize \\ $SSE_y = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
    \pause
    \vspace{\baselineskip}

    \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}{Parameter Estimation for CES Model Equation} %----------------------------
  \begin{center} \cesEqn \end{center}
  \begin{itemize}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \vspace{\baselineskip}

    \item{$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, $\rho$ estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Parameter Estimatino for LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Derived from thermodynamics \citep{Kummel:1980wx}}
    \item{Time-dependent technology $\rightarrow$ efficiency ratios}
    \item{Productivity ($A$) $\uparrow$ if}
    \begin{itemize}
      \item{Efficiency of producing capital stock increases ($k$ > $l + e$)}
      \item{Energy efficiency of labor increases ($l$ > $e$)}
    \end{itemize}
    \item{$a_0$ and $c_t$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}
